\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{semantic}
\usepackage{syntax}
\usepackage{verbatimbox}

\newcommand{\dom}{\operatorname{dom}}
\newcommand{\instantiates}{\operatorname{<:}}
\newcommand{\issatisfied}{\operatorname{is\, satisfied}}
\newcommand{\isfresh}{\operatorname{is\, fresh}}
\newcommand{\arefresh}{\operatorname{are\, fresh}}

\newcommand{\void}{\texttt{void}}

\newcommand{\hastype}[3]{\#1 \vdash #2 : #3}
\newcommand{\tuple}[1]{\left\langle #1 \right\rangle}
\newcommand{\yield}[2]{\tuple{#1} \rightarrow \tuple{#2}}

\title{The Smudge Language Definition}

\begin{document}

\maketitle

\section*{Grammar}
Smudge's syntax is comprised of top-level modules composed of state machine
definitions, in turn composed of one or more comma-separated transient- or
non-transient states, possibly intermixed with locally-scoped state machine
definitions.  Non-transient states are made up of zero or more enter and exit
side effects and one or more comma-separated event handlers, possibly
intermixed with locally-scoped state machine definitions.
%
Event handlers are either ``arrow'' (transitioning), ``dash''
(non-transitioning), or ``become'' (submachine-transitioning) declarations,
whose side effects can either be foreign functions or Smudge events, with or
without arguments delimited by spaces.
%
All identifiers can contain any character (when quoted) except foreign
functions, which are conservatively limited to C-style identifiers.

The rules are as follows:

\setlength{\grammarindent}{6em}
\begin{grammar}
\input{syntax.latex}
\end{grammar}

These productions are built out of the tokens extracted by the lexer.  These
include defined ASCII and Unicode character classes not reproduced here for the
sake of brevity:

\setlength{\grammarindent}{6em}
\begin{grammar}
\input{lexer.latex}
\end{grammar}

\section*{Type inference}

Smudge uses type inference both to determine the correctness of Smudge programs
and to facilitate code generation.
%
The type system Smudge uses is nominal, which is to say that machines with
different names are of different types, even if they handle the exact same set
of events, and differently named events are of different types even if they
contain the same data.
%
It is also effectful, which is to say that side effects do not return values,
but instead have associated effects (also called capabilities) when run.
%
Currently the only capability available is to handle the active event.

The types of terms in Smudge are implicit---there is no type ascription syntax.
%
There is also no way to explicitly define types---they are defined implicitly by
the syntax of machines and event handlers.
%
Accordingly, types within a single module are inferred using a deterministic
constraint solver.
%
Constraints are satisfied through monomorphic unification, save for effects and
capabilities, whose unification depends on the flags supplied at invocation.

To construct these rules, both the notation of natural deduction for type
inference and the notation for automata are useful here, with a slight
modification: rather than a machine being represented by a top-level structure
composed of states, symbols, and uniform total transitions, the symbols
(referred to as events) are nested behind the states, and nested within events
are transitions composed of a sequence of side-effectful computation
statements, whose types can be non-uniform.
%
For clarity, the following conventions are used:
\begin{itemize}
    \item $\Gamma$ refers to a type environment, an ordered mapping from names to types
    \item $\mathcal{S}$ refers to a signature environment, an ordered mapping from type names to their definition
    \item $\tau$ refers to a type
    \item $\alpha$ refers to a type variable (or ``tyvar'')
    \item $\phi$ refers to a capability or effect
    \item $C$ refers to the set of constraints
    \item $\theta$ refers to the substitutions
    \item $x$ refers to an identifier

    \item $m$ refers to a single state machine
    \item $q$ refers to a single state
    \item $a$ refers to a single input event
    \item $\delta$ refers to a total transition function for a state and event
    \item $\check{\delta}$ refers to a named side-effectful partial transition
    \item $e$ refers to an expression, which appear as subterms of a partial transition
\end{itemize}

Environments, types, and constraints are constructed according to the grammar
shown in Figure~\ref{fig:types}.

\begin{figure}
  \renewcommand{\syntleft}{\normalfont\itshape}
  \renewcommand{\syntright}{}
  \setlength{\grammarparsep}{2pt}
  \setlength{\grammarindent}{2em}
  \begin{grammar}
    <$\Gamma$,$\mathcal{S}$> ::= $\varnothing$ | \{ <x> $\mapsto \tau$, \ldots, <x> $\mapsto \tau$ \} | $\Gamma$ $\Gamma$

    <$\tau$> ::= $\alpha$ | "machine" <x> | $\phi$ | $\tau \rightarrow \tau$ | $\tau \times \tau$ | $\Pi\ \Gamma$ | $\Sigma\ \Gamma$

    <$\phi$> ::= $\varnothing$ | $\phi$, "eventful" <x>

    <C> ::= $\top$ | $\tau$ \textless: $\tau$ | $\tau \sim \tau$ | $\tau \sim$ "range" $\tau$ | <C> $\land$ <C>
  \end{grammar}
  \caption{\label{fig:types} Grammar of types, environments, and constraints}
\end{figure}

Environments are either the empty set $\varnothing$, designated as some
sequences of bindings of a name $x_i$ to type $\tau_i$, or constructed by
concatenating two environments $\Gamma_1 \Gamma_2$.

A type can either be a tyvar $\alpha$, some named \syntax{"machine" $x$}, an
effect/capability $\phi$, a function $\tau_p \rightarrow \tau_r$, a tuple (or
product) $\tau_1 \times \tau_2$, record $\Pi\ \Gamma$, or variant (also called
a sum or tagged union) $\Sigma\ \Gamma$.
%
Additionally, there is a restriction that the final type in a chain of arrows
(the return) must be $\phi$; currently functions do not chain (i.e., there are
no closures or partial application) and are expected to have one parameter that
is a tuple of the arguments, but that is not enforced in the type grammar.
%
A consequence of this is that side effects in Smudge cannot return values, so to
do work they have to send events.
%
Record types are equivalent to tuples with named entries, and so the order of
fields is fixed.
%
Variants on the other hand do not strictly police the order of their fields;
although Smudge does not include the unnamed sum type, this property is needed
for inference.

Effects or capabilities are either $\varnothing$ (no effect), or some set of
effects that includes \syntax{"eventful" $x$}, meaning that the active event of
type $x$ is available to the side effect in question.
%
When compiled, this will appear in C as a normal parameter to the function,
which returns \texttt{void}.

Constraints come in three flavors:
trivially true ($\top$),
subtype ($\tau_1 <: \tau_2$), and
equality ($\tau_1 \sim \tau_2$).
%
Multiple constraints are conjoined logically ($C_1 \land C_2$).
%
There is also a necessary operator \syntax{"range" $\tau$}, which represents the
label-erased tuple type for some record; that is, for some record type $\tau$
defined as $\Pi \{ x \mapsto \tau_1, y \mapsto \tau_2 \}$, the range is defined
to be $\tau_1 \times \tau_2$.

\subsection*{Constraint rules}

\begin{figure}
  \[\begin{array}{c}
    \inference[Module]
    {% Get the names of all of the machines
     m_1 = \textsc{machine}(x_1,\_) & \ldots & m_n = \textsc{machine}(x_n,\_) \\
     % Create the signature environment by mapping machine names to fresh type variables
     \mathcal{S}\equiv \{ x_i \mapsto \alpha_i \mid \alpha_i \isfresh, 1\leq i\leq n\} \\
     % Create the initial type environments by adding a singleton for each machine ...
     \Gamma_m\equiv \{ x_i \mapsto \texttt{machine } x_i \mid 1\leq i\leq n\} \\
     % ... and scraping the entire program for foreign function names
     \Gamma_\delta\equiv \{ x \mapsto \alpha \mid \alpha \isfresh, \textsc{function}(x) \in m_1,\ldots,m_n \} \\
     % Recurse, typing each machine
     C_1,\mathcal{S},\Gamma \Gamma_\delta \Gamma_m \vdash m_1 : \tau_1 & \ldots
     & C_n,\mathcal{S},\Gamma \Gamma_\delta \Gamma_m \vdash m_n : \tau_n \\
     % Combine all constraint clauses and equate the returned types
     C\equiv \bigwedge_{i=1}^n C_i \land \alpha_i \sim \tau_i &
     % Solve the constraint system
     \theta C\issatisfied}
    {\top,\varnothing,\Gamma \vdash \textsc{module}(\tuple{m_1,\ldots,m_n}) : \Pi\ \theta\mathcal{S}}
    \\\\
    \inference[StateMachine]
    {% Recurse, typing each state
     C_1,\mathcal{S},\Gamma \vdash q_1 : \tau_1 & \ldots & C_n,\mathcal{S},\Gamma \vdash q_n : \tau_n \\
     % Combine all constraint clauses and build the supertype of the returned state types
     C\equiv \bigwedge_{i=1}^n C_i \land \tau_i <: \alpha & \alpha \isfresh}
    {C,\mathcal{S},\Gamma \vdash \textsc{machine}(\_,\tuple{q_1,\ldots,q_n}) : \alpha}
    \\\\
    \inference[State]
    {% Get the names of all of the events
     a_1 = \textsc{event}(x_1,\_,\_) & \ldots & a_n = \textsc{event}(x_n,\_,\_) \\
     % Recurse, typing each event handler
     C_1,\mathcal{S},\Gamma \vdash a_1 : \tau_1 & \ldots & C_n,\mathcal{S},\Gamma \vdash a_n : \tau_n \\
     % Combine all constraint clauses
     C\equiv \bigwedge_{i=1}^n C_i &
     % Create the state's type environment by mapping event names to their sender
     \Gamma_q\equiv \{ x_i \mapsto \tau_i \mid 1\leq i\leq n\} }
    {C,\mathcal{S},\Gamma \vdash \textsc{state}(\_,\tuple{a_1,\ldots,a_n}) : \Sigma\ \Gamma_q}
    \\\\
    \inference[Event]
    {% Create the event's type environment by mapping field names to fresh tyvars
     \Gamma_a\equiv \{ x_i \mapsto \alpha_i \mid \alpha_i \isfresh, 1\leq i\leq m\} \\
     % Recurse, typing each side effect
     C_1,\mathcal{S},\Gamma \vdash \check{\delta}_1 : \phi_1 & \ldots & C_n,\mathcal{S},\Gamma \vdash \check{\delta}_n : \phi_n \\
     % The side effects all return an effect, which could include eventful x
     C\equiv \bigwedge_{i=1}^n C_i \land \texttt{eventful } x <: \phi_i}
    {C,\mathcal{S},\Gamma \vdash \textsc{event}(x,\tuple{x_1,\ldots,x_m},\tuple{\check{\delta}_1,\ldots,\check{\delta}_n}) : \Pi\ \Gamma_a \rightarrow \varnothing}
    \\\\
    \inference[Effect]
    {% Recurse to type the effect function, and make a tyvar for the effect
     C_{\check{\delta}},\Gamma \vdash e_{\check{\delta}} : \tau_{\check{\delta}} & \alpha_\phi \isfresh \\
     % Recurse to type the effect arguments
     C_1,\mathcal{S},\Gamma \vdash e_1 : \tau_1 & \ldots & C_n,\mathcal{S},\Gamma \vdash e_n : \tau_n \\
     % Constrain the function to be a function of arity n of the corresponding type and effect, then combine all constraint clauses
     C\equiv \tau_{\check{\delta}} \sim \tau_1 \times \ldots \times \tau_n \rightarrow \alpha_\phi \land C_{\check{\delta}} \land \bigwedge_{i=1}^n C_i}
    {C,\mathcal{S},\Gamma \vdash \textsc{effect}(e_{\check{\delta}},\tuple{e_1,\ldots,e_n}) : \alpha_\phi}
    \\\\
    \inference[SendEvent]
    {% Look up the machine's signature, and create three fresh tyvars
     \tau = \mathcal{S}(x_m) & \alpha_a,\alpha_\pi,\alpha_\phi \arefresh \\
     % Constrain the signature to include a field of this name with a function type, constrain the argument to be a record, and break out the range of that record
     C\equiv \Sigma\ \{ x_a \mapsto \alpha_a \rightarrow \varnothing \} <: \tau \land \Pi\ \varnothing <: \alpha_a \land \alpha_\pi \sim \texttt{range } \alpha_a}
    {C,\mathcal{S},\Gamma \vdash \textsc{sendevent}(x_m, x_a) : \alpha_\pi \rightarrow \alpha_\phi}
    \\\\
    \inference[Function]
    {% Look up the name of the variable
     \tau = \Gamma(x)}
    {C,\mathcal{S},\Gamma \vdash \textsc{function}(x) : \tau}
    \\\\
    \inference[Label]
    {% Look up the name of the variable
     \tau = \Gamma(x)}
    {C,\mathcal{S},\Gamma \vdash \textsc{label}(x) : \tau}
  \end{array}\]
  \caption{\label{fig:inference} Type inference}
\end{figure}

Elaboration consists of three phases: constraint inference, unification, and
instantiation.
%
We start with a discussion of inference.

Figure~\ref{fig:inference} shows the rules of Smudge's type system.
%
The judgment form $C,\mathcal{S},\Gamma \vdash e : \tau$ means given
constraints $C$, machine signatures $\mathcal{S}$, and type environment
$\Gamma$, term $e$ is well formed and has type $\tau$.

The algorithm begins at the module level.
%
The Module rule starts with the built-in type environment $\Gamma$, but an empty
signature environment, as Smudge does not currently have any built-in machines.
%
It pulls the names $x_i$ out of each machine definition, and creates the
module's signature environment $\mathcal{S}$ by binding each name to a fresh
tyvar.
%
It then creates the type environment $\Gamma_m$, which includes the singleton
for each machine, and scrapes the entire module for foreign function names,
which it puts in the type environment $\Gamma_\delta$.
%
Next, inference recurses, passing in the signature environment and combined type
environments to find the type $\tau_i$ and constraints $C_i$ for each machine
$m_i$.
%
These constraints are conjoined, with the additional constraints that the tyvars
reserved for our machines must match the types inferred for them.
%
These constraints are then satisfied by the unification algorithm, which yields
the substitution $\theta$.
%
The type of the overall module is found by applying this substitution to the
signature environment, then compiling it into a record $\Pi\ \theta\mathcal{S}$

The StateMachine rule simply recurses, typing each state $q_i$ individually.
%
It creates a fresh tyvar $\alpha$ to use as the type of the machine, and
constrains it to be the supertype over all states, then combines these with all
the other constraint clauses.

The State rule begins by pulling the names $x_i$ out of each event handler $a_i$.
%
It recurses, typing each handler, and combines all the constraints.
%
It then creates the state's type environment $\Gamma_q$ by mapping event name
$x_i$ to the type $\tau_i$ inferred for the corresponding handler.

To infer the type for event handlers, the Event rule creates the event's type
environment $\Gamma_a$ by mapping field names $x_1$ through $x_m$ to fresh
tyvars.
%
It next recurses to the side effects $\check{\delta}$.
%
Because the return type for all side effects is an effect/capability, we use
$\phi$ to make this arrangement explicit.
%
We conjoin all constraints, adding additional subtype constraints which
augment each $\phi_i$ with the capability \syntax{"eventful" $x$}.
%
The handler is then given the odd-looking type
$\Pi\ \Gamma_a \rightarrow \varnothing$.
%
This is a function that takes a record over all of the handler's fields and has
no effects.
%
There are two noteworthy things here---one, that this does not introduce a named
record type into the signature environment as might be expected, relying instead
on the definition of the machine; and two, that the type takes a record instead
of a product, which is out of step with the concrete syntax.
%
We address that in the SendEvent rule.

The Effect rule recurses to type both the function expression
$e_{\check{\delta}}$ and the argument expressions $e_i$ (we call these
expressions for lack of a better term---in truth, they currently just range over
names of events, foreign functions, and handler fields).
%
With the help of the effect tyvar $\alpha_\phi$, it then constrains the function
type $\tau_{\check{\delta}}$ to be of arity $n$ of the corresponding argument
types, then combines this with all constraint clauses.
%
As each side effect represents a complete application, the returned type is the
effect tyvar $\alpha_\phi$.

When the expression in question is a fully qualified event, the SendEvent rule
looks up the signature $\tau$ of the machine $x_m$ in $\mathcal{S}$, and creates
three fresh tyvars: $\alpha_a$ for the event record, $\alpha_\pi$ for the
label-erased tuple type, and $\alpha_\phi$ for the effect.
%
To type correctly, the rule performs a few tricks.
%
First, it constrains the machine's signature to include a binding for the event
$x_a$ of function type.
%
Second, it constrains the domain of that function to have a record type, but
does not specify any field names.
%
Third, it adds a constraint that erases the labels for that record type, taking
only the range as the product type $\alpha_\pi$.
%
During code generation, the external API provides only the version that accepts
a complete record; the tuple version is only available internally.
%
With the help of these tricks, the returned type is now compatible with the
constraint from the Effect rule.

Finally, the Function and Label rules are straightforward.
%
Both look up the type for the name in the type environment and return it.

\subsection*{Unification}

\begin{figure}
  \[\begin{array}{c}
    \inference[Conjoin]
    {\theta_1 C_1\issatisfied&\theta_2 (\theta_1 C_2)\issatisfied}
    {(\theta_2 \circ \theta_1)C_1\land C_2\issatisfied}
    \\\\
    \inference[Trivial]
    {}
    {\top\issatisfied}
    \\\\
    \inference[EqualReflexive]
    {}
    {\tau\sim \tau\issatisfied}
    \\\\
    \inference[EqualTyvar]
    {\alpha\not\in \textrm{ftv}(\tau) &
     \theta\equiv \{\alpha \mapsto \tau \}}
    {\theta(\alpha\sim \tau)\issatisfied}
    \\\\
    \inference[EqualCapability]
    {\theta (\phi_1, \phi_2 \sim \phi_3, \phi_4)\issatisfied}
    {\theta(\phi_1, \texttt{eventful }x, \phi_2 \sim \phi_3, \texttt{eventful }x, \phi_4)\issatisfied}
    \\\\
    \inference[EqualArrow]
    {\theta (\tau_1 \sim \tau_3 \land \tau_2 \sim \tau_4)\issatisfied}
    {\theta(\tau_1 \rightarrow \tau_2 \sim \tau_3 \rightarrow \tau_4)\issatisfied}
    \\\\
    \inference[EqualProduct]
    {\theta (\tau_1 \sim \tau_3 \land \tau_2 \sim \tau_4)\issatisfied}
    {\theta(\tau_1 \times \tau_2 \sim \tau_3 \times \tau_4)\issatisfied}
    \\\\
    \inference[EqualRecord]
    {\Gamma_x\equiv\{ x_1 \mapsto \tau_{x_1}, \ldots, x_m \mapsto \tau_{x_m} \} \\
     \Gamma_y\equiv\{ y_1 \mapsto \tau_{y_1}, \ldots, y_n \mapsto \tau_{y_n} \} \\
     m = n & x_1 = y_1 \ldots x_m = y_n \\
     \theta (\tau_{x_1} \sim \tau_{y_1} \land \ldots \land \tau_{x_m} \sim \tau_{y_n})\issatisfied}
    {\theta(\Pi\ \Gamma_x\sim \Pi\ \Gamma_y)\issatisfied}
    \\\\
    \inference[EqualVariant]
    {\Gamma_x\equiv\{ x_1 \mapsto \tau_{x_1}, \ldots, x_m \mapsto \tau_{x_m} \} \\
     \Gamma_y\equiv\{ y_1 \mapsto \tau_{y_1}, \ldots, y_n \mapsto \tau_{y_n} \} \\
     m = n &
     \theta (\bigwedge_{i=1}^m \tau_{x_i} \sim \tau_{y_j} \mid x_i = y_j)\issatisfied}
    {\theta(\Sigma\ \Gamma_x\sim \Sigma\ \Gamma_y)\issatisfied}
    \\\\
    \inference[EqualRange]
    {\Gamma_y\equiv\{ y_1 \mapsto \tau_{y_1}, \ldots, y_n \mapsto \tau_{y_n} \} \\
     m = n &
     \theta (\tau_{x_1} \sim \tau_{y_1} \land \ldots \land \tau_{x_m} \sim \tau_{y_n})\issatisfied}
    {\theta(\tau_{x_1} \times \ldots \times \tau_{x_m}\sim \texttt{range } \Pi\ \Gamma_y)\issatisfied}
  \end{array}\]
  \caption{\label{fig:unification} Constraint unification for non-subtype constraints}
\end{figure}

After accruing constraints within a module, unification satisfies them, yielding
an assignment of tyvars to types, commonly known as a substitution.
%
Figure~\ref{fig:unification} gives the rules used to solve conjunctions of
trivially true and equality constraints, and the subtyping rules are listed in
Figure~\ref{fig:unificationsubtype}.
%
Let us first discuss the non-subtype rules.

To solve multiple constraints, the Conjoin rule first finds a substitution
$\theta_1$ that satisfies the first constraint $C_1$, then applies that
substitution to $C_2$ before solving it to get $\theta_2$.
%
The resulting substitution that solves the conjunction is the composition of the
two, $\theta_2\circ\theta_1$.
%
We must solve these in order rather than independently to ensure a valid
solution; solving them independently may result in an error.

The Trivial rule covers trivially true constraints, which are always satisfied.

One of the simplest rules is EqualReflexive, which states that a type must be
equal to itself.

The core of unification is the EqualTyvar rule, which is the source of
substitutions.
%
It creates a binding of the tyvar $\alpha$ to the arbitrary type $\tau$, with
the critical caveat that $\alpha$ does not appear in the free tyvars of $\tau$.

As capabilities form a set, the order does not matter.
%
Accordingly, the EqualCapability rule sidesteps the exact order of capabilities;
provided that both sets contain the \syntax{"eventful" $x$} capability and the
remaining sets match, the constraint is satisfied.

For typing functions, we use the EqualArrow rule.
%
This checks that the parameters for both functions, $\tau_1$ and $\tau_3$,
match, and the same for the results $\tau_2$ and $\tau_4$.

The EqualProduct rule for tuples looks similar and behaves the same.

The EqualRecord rule sooks intimidating, but is actually straightforward.
%
It starts by splitting out both records to expose the field names and their
types.
%
It then checks that both records have the same number of fields, and that the
field names of $\Gamma_x$ are the same as those of $\Gamma_y$---two records with
the same field names in a different order are \emph{not} the same type.
%
To derive the substitution $\theta$, it runs unification on the conjunction of
the equality of the fields' types.

The EqualVariant is similar, but with the twist that the field name order does
not matter---two variants with the same field names in different order compare
the same, and are thus the same type (so long as no fields are missing).
%
We achieve this by iterating the fields of $\Gamma_x$, finding the same field
from $\Gamma_y$, and unifying their types.

Finally, the EqualRange rule is a twist on EqualRecord---this time, we need only
break apart $\Gamma_y$.
%
As we are comparing it to a product type, we can ignore the field names, and
just unify the conjunction over the types.
%
It is important to note here that this constraint is harder to solve, as it only
accepts a product and a record; as such, these constraints have to be deferred
until the end of unification to guarantee that they can be solved.

\begin{figure}
  \[\begin{array}{c}
    \inference[SubtypeReflexive]
    {}
    {\tau <: \tau\issatisfied}
    \\\\
  \end{array}\]
  \caption{\label{fig:unificationsubtype} Constraint unification for subtype constraints}
\end{figure}

\subsection*{Monomorphic unification rules}
These rules dictate type resolution in the case where the types must match.
This behavior can be selected by supplying a flag at invocation.  It resolves
conflicting constraints by rejecting the module as ill-formed.

\[
    \inference[SubtypeEqual]
    {\tau_1=\tau_2}{\tau_1\instantiates\tau_2}
\]

\subsection*{Subtype polymorphic unification rules}
The default unification method is to simplify the types using polymorphism.  It
is instructive to first consider a non-deterministic rule for simplification.
According to this rule, a simplified type is a type that is strictly more
general than either of two other types, such that these two types are subtypes
of it.
\[
    \inference[SimplifyNonDeterministic]
    {\hastype{\Gamma}{\check{\delta}_1}{\tau_1\times\ldots\times\tau_n\rightarrow\tau}\\
     \hastype{\Gamma}{\check{\delta}_2}{\tau_1'\times\ldots\times\tau_m'\rightarrow\tau'}\\
     \tau_1\times\ldots\times\tau_n\rightarrow\tau\instantiates\tau_1''\times\ldots\times\tau_k''\rightarrow\tau''\\
     \tau_1'\times\ldots\times\tau_m'\rightarrow\tau'\instantiates\tau_1''\times\ldots\times\tau_k''\rightarrow\tau''}
    {\hastype{\Gamma}{\textsc{simplify}(\check{\delta}_1, \check{\delta}_2)}{\tau_1''\times\ldots\times\tau_k''\rightarrow\tau''}}
\]

However, this definition is insufficient to specify the actual result of type
unification.  For this, the accumulated constraints must be resolved into a
supertype through additional rules that build on top of the monomorphic ones.

These include the basic subtyping rules that dictate structural relationships
directly between types.  For clarity the n-ary rule is included, although it is
implicit in the arrow rule.

A key detail worth drawing attention to in these rules is that function return
and parameter types are generalized in the opposite direction --- return types
covariantly, and parameter types contravariantly.  This is because for a
function, a concrete return type is actually a more general type.

\[
    \inference[SubtypeVoid]
    {}{\void\instantiates\tau}
\]

\[
    \inference[SubtypeReflexive]
    {}{\tau\instantiates\tau}
\]

\[
    \inference[SubtypeArrow]
    {\tau''\instantiates\tau&\tau'\instantiates\tau'''}
    {\tau\rightarrow\tau'\instantiates\tau''\rightarrow\tau'''}
\]

\[
    \inference[SubtypeN-ary]
    {\tau_1'\instantiates\tau_1&\tau_2\times\ldots\times\tau_n\rightarrow\tau\instantiates\tau_2'\times\ldots\times\tau_n'\rightarrow\tau'}
    {\tau_1\times\tau_2\times\ldots\times\tau_n\rightarrow\tau\instantiates\tau_1'\times\tau_2'\times\ldots\times\tau_n'\rightarrow\tau'}
\]

In addition, meet and join rules dictate relationships that are use-dependent.
Although Smudge is only capable of directly inferring joins, meet rules are
required to resolve parameter contravariance.

\[
    \inference[SubtypeMeetLeft]
    {}{\tau_1\land\tau_2\instantiates\tau_1}
\]

\[
    \inference[SubtypeMeetRight]
    {}{\tau_1\land\tau_2\instantiates\tau_2}
\]

\[
    \inference[SubtypeMeet]
    {\tau\instantiates\tau_1&\tau\instantiates\tau_2}
    {\tau\instantiates\tau_1\land\tau_2}
\]

\[
    \inference[SubtypeJoinLeft]
    {}{\tau_1\instantiates\tau_1\lor\tau_2}
\]

\[
    \inference[SubtypeJoinRight]
    {}{\tau_2\instantiates\tau_1\lor\tau_2}
\]

\[
    \inference[SubtypeJoin]
    {\tau_1\instantiates\tau&\tau_2\instantiates\tau}
    {\tau_1\lor\tau_2\instantiates\tau}
\]

\[
    \inference[SubtypeJoinArrow]
    {}
    {\tau\rightarrow\tau'\lor\tau''\rightarrow\tau'''\instantiates
     (\tau''\land\tau)\rightarrow(\tau'\lor\tau''')}
\]

\section*{Foreign function interface}

Access to the host language from Smudge is afforded by calls to
foreign functions.  The interfaces of such functions are derived from
the context in which they are invoked, subject to type unification
rules, and the definition is provided by the programmer.  The
interface is uniform per backend, and the semantics of the call are
common to all backends.

Access to Smudge from the host language is afforded by calls to event
functions.  Such functions accept a parameter holding the data
associated with this event, and the definition is generated by the
compiler.

\subsection*{Foreign functions in the C backend}

Event struct typedefs emitted by the compiler have names generated
according to the name mangling translation.

Foreign function declarations emitted by the compiler have the exact
name stated in the source, accept zero or more parameters, and return
nothing.
%
All function parameters have function pointer type, except for an
optional initial const pointer to an event struct.

Event function declarations emitted by the compiler have names generated
according to the name mangling translation and accept one parameter,
a const pointer to an event struct.

A foreign function invoked only in the handler for a single type of
named event will always accept an initial parameter of the corresponding
type.
%
One invoked only as an enter or exit side effect will not receive this
initial parameter.
%
One invoked from multiple contexts---either the
handlers for multiple named events or from both named event handlers
and enter or exit side effects---loses this parameter under
polymorphic unification, but causes a type error under monomorphic
unification.

All other parameters are as passed in the source file.
%
The types of these parameters are inferred from their use sites, where
functions take either zero or one parameters, according to the
preceding rules, and events have the type of the corresponding
generated event-sending function.

\section*{Name mangling}

For most backends, the input alphabet of Smudge identifiers is larger than the
output alphabet because Smudge identifiers can include, e.g., non-alphanumeric
symbols and whitespace.

As a consequence, those backends are required to somehow translate identifiers
into their output alphabet.  This process is known as name mangling.

Because of the nature of identifiers, an important goal for any name mangling
translation is to maintain their uniqueness: it should be injective, i.e., no
two qualified input identifiers should yield the same output identifier.  The
basis for this is intuitive: static checks are run on unmangled identifiers,
not mangled identifiers.  If two Smudge identifiers mangle to the same output,
the correctness of those checks is not preserved; this will lead to errors in
the output.  While some of them will be caught by static checks in the target
compiler, others could result in runtime errors.

Because the mangled names generated by Smudge have to be used externally, it
also makes sense to try to mangle the minimum possible while maintaining
injectivity.

\subsection*{Name mangling in the C backend}

Identifiers in C are the language \verb|[_a-zA-Z][_a-zA-Z0-9]*|.

To map into this space as densely as possible, unqualified Smudge identifiers
are handled as three different cases:

\begin{enumerate}
    \item \verb|[a-zA-Z0-9]+| \\
        The mapping is identity.
    \item \verb|[a-zA-Z0-9]+(_[a-zA-Z0-9]+)+| \\
        The mapping is identity followed by a trailing underscore.
    \item \verb|.*| \\
        The mapping is to append an underscore, the decimal length, and two
        underscores, i.e., \verb|_len__|, and mangle each character as follows:
        \begin{enumerate}
            \item \verb|[a-zA-Z0-9]| \\
                The mapping is identity.
            \item \verb|_| \\
                The mapping is double underscore.
            \item \verb|.| \\
                The mapping is the UTF-8 hexadecimal encoding of the character
                wrapped in leading and trailing underscores.
        \end{enumerate}
\end{enumerate}

Note that the output identifier language is \verb|[_a-zA-Z0-9]+| rather than
C's; i.e., this mapping can produce output with leading underscores and decimal
digits; the former is undefined behavior in C, while the latter is not in the
language of C identifiers.  Leading digits are repaired below during qualified
identifier translation.  Leading underscores are handled elsewhere, and no
further consideration is given to them here.

Qualified identifiers are mangled by mangling unqualified identifiers, and then
joining them by concatenation with two underscores if the second unqualified
identifier is composed of alphanumeric characters separated by lone underscores,
and concatenation with a single underscore otherwise.

In order to prove the injectivity of the overall translation, the unqualified,
qualified, and single character translations can be separately proven
injective.  Because the overall translation is a composition of these, if they
are individually injective then the composition is injective.  An informal
proof follows.

The single character translation uses identity on alphanumeric characters,
which is injective with respect to itself.  Underscores are doubled, which is
trivially injective with respect to itself, and which cannot produce an
alphanumeric character, and alphanumeric character identity cannot produce two
underscores, so these are injective with respect to each other.  Otherwise, the
character is translated to a hexadecimal number wrapped by underscores; this
number is assumed to be an injective mapping defined by UTF-8, and because it
is wrapped by underscores, it cannot produce a lone alphanumeric character, and
since all numbers are at least one digit long and contain no underscores, it
cannot produce an output consisting solely of underscores, and thus it is
injective with respect to itself as well as the other two cases; finally,
alphanumeric character identity cannot produce underscores, and doubling
underscores cannot produce hexadecimal numbers.  All cases are injective with
respect to themselves and each other, and so single character translation is
injective.

The concatenation of the output of the single character translation can be
proven injective by induction on the length of the input.  The single character
base case is presented previously.  A total function with only positive-length
outputs, the empty string is thus also injective.  The inductive hypothesis is
that all inputs $b_1\ldots b_k$ of length $k$ are injective with respect to all
inputs $a_1\ldots a_j$ of length $j\leq k$ .  The proof of the induction is by
contradiction: assume there are two such inputs $a\neq b$ of length $j+1$ and
$k+1$ respectively s.t.
\begin{align*}
    t(a_1)+\ldots+t(a_j)+t(a_{j+1})&=    t(b_1)+\ldots+t(b_k)+t(b_{k+1})\\
    t(a_1)+\ldots+t(a_j)           &\neq t(b_1)+\ldots+t(b_k)
\end{align*}
where $t()$ denotes single character translation and plus denotes
concatenation.  In order for this to be the case, either $t(a_1)+\ldots+t(a_j)$
or $t(b_1)+\ldots+t(b_k)$ must be a proper prefix of the other\footnote{It
doesn't matter which, because the same arguments would apply.}; say it is
$t(a_1)+\ldots+t(a_j)$.  As such, $t(a_{j+1})$ is a proper suffix of
$t(b_1)+\ldots+t(b_{k+1})$, which means more simply that either $t(a_{j+1})$ or
$t(b_{k+1})$ is a suffix of the other.  Consider the cases of the single
character translation: alphanumeric identity, underscore doubling, and UTF-8
underscore wrapping produce outputs of length 1, 2, and at least 3,
respectively, which means that the latter cannot suffix the former;
alphanumeric identity cannot produce underscores, and so cannot be a suffix of
the other cases; finally, UTF-8 underscore wrapping has a lone underscore
suffix, no more, while underscore doubling cannot produce a lone underscore,
only more.  Thus neither $t(a_{j+1})$ nor $t(b_{k+1})$ can be a proper suffix
of the other, so they must be equal.  As as $t(a_{j+1})=t(b_{k+1})$, trimming
them off of their respective outputs gives
\[t(a_1)+\ldots+t(a_j)=t(b_1)+\ldots+t(b_k)\]
which contradicts the premise that they differ.  Thus, $a=b$ and the
concatenation for lengths $j+1, k+1$ is injective for all $k,j\leq k$.
Concatenation is associative, so this is true for prepending as well.

The unqualified identifier translation uses identity on alphanumeric strings
(the ``alphanumeric'' case), which is injective with respect to itself.
Strings composed of alphanumeric characters separated by lone underscores (the
``single underscore'' case) are translated using identity followed by a
trailing underscore, which is injective with respect to itself because identity
and concatenation with a fixed pattern are both injective, and since it cannot
produce purely alphanumeric output and the alphanumeric case cannot produce a
trailing underscore, these cases are injective with respect to each other.
Otherwise (the ``general'' case), the translation is the concatenation of the
single character translation applied to the input, presented previously,
followed by an underscore followed by the input length encoded in decimal
followed by two underscores; decimal natural number encoding is injective, as
is concatenation with the stated fixed underscore pattern, and because
concatenation of the single character translation is injective there are no two
inputs of the same lengths that produce the same output, and as it produces
output with two trailing underscores, it cannot produce output with a trailing
alphanumeric character or solitary underscore, so overall this translation is
injective with respect to itself as well as the other two cases; finally,
neither the alphanumeric nor single underscore case can produce two trailing
underscores.  All cases are injective with respect to themselves and each
other, and so unqualified identifier translation is injective.

The qualified identifier translation, a concatenation of the output of the
unqualified identifier translation intercalated with underscores, can be proven
injective by induction on the length of the input.  The single identifier base
case is presented previously.  A total function with only non-negative-length
outputs, the empty qualified identifier is thus also injective.  The inductive
hypothesis is that all inputs $y_1\ldots y_k$ of length $k$ are injective with
respect to all inputs $x_1\ldots x_j$ of length $j\leq k$.  The proof of the
induction is by contradiction: assume there are two such inputs $x\neq y$ of
length $j+1$ and $k+1$ respectively s.t.
\begin{myverbbox}{\vunder}_\end{myverbbox}
\begin{align*}
    u(x_1)+\ldots+\vunder+u(x_j)+\vunder+u(x_{j+1})&=    u(y_1)+\ldots+\vunder+u(y_k)+\vunder+u(y_{k+1})\\
    u(x_1)+\ldots+\vunder+u(x_j)                   &\neq u(y_1)+\ldots+\vunder+u(y_k)
\end{align*}
where $u()$ denotes unqualified identifier translation, underscores denote the
join sequence\footnote{See previous specification for qualified identifier
mangling.}, and plus denotes concatenation.  In order for this to be the case,
either \mbox{$u(x_1)+\ldots+\vunder+u(x_j)$} or $u(y_1)+\ldots+\vunder+u(y_k)$
must be a proper prefix of the other\footnote{Again, it doesn't matter which};
say it is $u(x_1)+\ldots+\vunder+u(x_j)$.  As such, $\vunder+u(x_{j+1})$ is a
proper suffix of $u(y_1)+\ldots+\vunder+u(y_{k+1})$, which means more simply
that either $\vunder+u(x_{j+1})$ or $\vunder+u(y_{k+1})$ is a suffix of the
other\footnote{Without the intercalated underscores, trivial examples of output
suffixing can easily be produced; as such, the rest of the proof requires the
intercalated underscores.}.  Consider the cases of the unqualified identifier
translation: the alphanumeric case, the single underscore case, and the general
case, each with underscore(s) prepended.  The alphanumeric case output cannot
have a trailing underscore so it cannot be a suffix of the other two cases.
The prepended underscore prevents its output from being a proper suffix of
other alphanumeric case outputs, as underscore is not in the output alphabet.
The single underscore case ends with exactly one trailing underscore so it
cannot be a suffix of the other two cases.  Two prepended underscores prevent
its output from being a proper suffix of other single underscore case outputs,
which cannot contain a sequence of two underscores.  The general case ends with
two underscores so it cannot be a suffix of the other two cases.  With respect
to itself, the prepended underscore does not help the general case; instead,
the length code is used.  The length suffix is injective with respect to the
length of the input identifier, thus if $\vunder+u(x_{j+1})$ or
$\vunder+u(y_{k+1})$ suffixes the other, $x_{j+1}$ and $y_{k+1}$ must have the
same length.  If the translation of a suffix of either input could equal the
translation of the entirety of the other, this would be a violation of the
injectivity already proven in the unqualified case.  Thus neither
$\vunder+u(x_{j+1})$ nor $\vunder+u(y_{k+1})$ can be a proper suffix of the
other, so they must be equal.  As as $\vunder+u(x_{j+1})=\vunder+u(y_{k+1})$,
trimming them off of their respective outputs gives
\[u(x_1)+\ldots+\vunder+u(x_j)=u(y_1)+\ldots+\vunder+u(y_k)\]
which contradicts the premise that they differ.  Thus, $x=y$ and the
concatenation for lengths $j+1, k+1$ is injective for all $k,j\leq k$.
Concatenation is associative, so this is true for prepending as
well\footnote{Although the proof in that direction is more convoluted, because
all cases can output proper prefixes of some general case output.}.

When the qualified identifier translation of an input $x$ produces an output
$q(x)$ with a leading decimal digit, an additional qualified join sequence is
performed, $\texttt{""}+\vunder+q(x)$.  The unqualified identifier translation
is incapable of yielding the empty string, and so the proof reduces to that of
the qualified identifier translation.  For illustrative purposes, though,
consider some $y$ s.t. $q(y)=\texttt{""}+\vunder+q(x)$: with leading
underscore(s), $y_1$ cannot be the alphanumeric or single underscore cases; it
and $x_1$ must be the general case.  However, if that is true, $q(x)$ cannot be
a suffix of $q(y)$, because the length codes must differ.

With the unqualified, qualified, and single character translations separately
proven injective, and as the overall translation is a composition of these, the
overall translation is thus proven injective.\qed

\end{document}
